---
title: "Uber Data Analysis"
output: html_notebook
---



**In this project, we will analyze the Uber Pickups in New York City dataset.**

### 1. Importing packages

```{r}
library(ggplot2)
library(ggthemes)
library(lubridate)
library(dplyr)
library(tidyr)
library(DT)
library(scales)
```


### 2. Creating vector of colors to be implemented in plots

```{r}
colors = c("#CC1011", "#665555", "#05a399", "#cfcaca", "#f5e840", "#0683c9", "#e075b0")
```


### 3. Reading the Data into their designated variables

**we will read several csv files that contain the data from April 2014 to September 2014. We will store these in corresponding data frames like apr_data, may_data, etc. After we have read the files, we will combine all of this data into a single dataframe called 'data_2014'.**

```{r}
unzip("./data/uber-raw-data-apr14.csv.zip", exdir="./data")
unzip("./data/uber-raw-data-may14.csv.zip", exdir="./data")
unzip("./data/uber-raw-data-jun14.csv.zip", exdir="./data")
unzip("./data/uber-raw-data-jul14.csv.zip", exdir="./data")
unzip("./data/uber-raw-data-aug14.csv.zip", exdir="./data")
unzip("./data/uber-raw-data-sep14.csv.zip", exdir="./data")
```

```{r}
fn <- "./data/uber-raw-data-apr14.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/uber-raw-data-may14.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/uber-raw-data-jun14.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/uber-raw-data-jul14.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/uber-raw-data-aug14.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/uber-raw-data-sep14.csv.zip"
if (file.exists(fn)) file.remove(fn)
```

```{r}
uber_apr <- read.csv(file = "./data/uber-raw-data-apr14.csv")

uber_may <- read.csv(file = "./data/uber-raw-data-may14.csv")

uber_jun <- read.csv(file = "./data/uber-raw-data-jun14.csv")

uber_jul <- read.csv(file = "./data/uber-raw-data-jul14.csv")

uber_aug <- read.csv(file = "./data/uber-raw-data-aug14.csv")

uber_sep <- read.csv(file = "./data/uber-raw-data-sep14.csv")
```

```{r}
uber_2014 <- rbind(uber_apr, uber_may, uber_jun, uber_jul, uber_aug, uber_sep)
```

```{r}
uber_2014$Date.Time <- as.POSIXct(uber_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S")
uber_2014$Time <- format(as.POSIXct(uber_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S"), format="%H:%M:%S")
uber_2014$Date.Time <- ymd_hms(uber_2014$Date.Time)
uber_2014$day <- factor(day(uber_2014$Date.Time))
uber_2014$month <- factor(month(uber_2014$Date.Time, label = TRUE))
uber_2014$year <- factor(year(uber_2014$Date.Time))
uber_2014$dayofweek <- factor(wday(uber_2014$Date.Time, label = TRUE))
uber_2014$hour <- factor(hour(hms(uber_2014$Time)))
uber_2014$minute <- factor(minute(hms(uber_2014$Time)))
uber_2014$second <- factor(second(hms(uber_2014$Time)))
```


#### Plotting the trips by the hours in a day

**we will use the ggplot function to plot the number of trips that the passengers had made in a day. We will also use dplyr to aggregate our data. In the resulting visualizations, we can understand how the number of passengers fares throughout the day. We observe that the number of trips are higher in the evening around 5:00 and 6:00 PM.**

```{r}
hour_data <- uber_2014 %>%
           group_by(hour) %>%
               dplyr::summarize(Total = n()) 
datatable(hour_data)
```

```{r}
ggplot(hour_data, aes(hour, Total)) + 
        geom_bar( stat = "identity", fill = "steelblue", color = "red") +
           ggtitle("Trips Every Hour") +
            theme(legend.position = "none") +
            scale_y_continuous(labels = comma)


month_hour <- uber_2014 %>%
          group_by(month, hour) %>%
             dplyr::summarize(Total = n())
ggplot(month_hour, aes(hour, Total, fill = month)) + 
       geom_bar( stat = "identity") +
          ggtitle("Trips by Hour and Month") +
           scale_y_continuous(labels = comma)
```


#### Plotting data by trips during every day of the month

**We observe from the resulting visualization that 30th of the month had the highest trips in the year which is mostly contributed by the month of April.**

```{r}
day_group <- uber_2014 %>%
          group_by(day) %>%
             dplyr::summarize(Total = n()) 
datatable(day_group)
```

```{r}
ggplot(day_group, aes(day, Total)) + 
        geom_bar( stat = "identity", fill = "steelblue") +
           ggtitle("Trips Every Day") +
            theme(legend.position = "none") +
            scale_y_continuous(labels = comma)
```

```{r}
day_month_group <- uber_2014 %>%
         group_by(month, day) %>%
             dplyr::summarize(Total = n())
ggplot(day_month_group, aes(day, Total, fill = month)) + 
        geom_bar( stat = "identity") +
           ggtitle("Trips by Day and Month") +
            scale_y_continuous(labels = comma) +
            scale_fill_manual(values = colors)
```


#### Number of Trips taking place during months in a year

**In this section, we will visualize the number of trips that are taking place each month of the year. In the output visualization, we observe that most trips were made during the month of September. Furthermore, we also obtain visual reports of the number of trips that were made on every day of the week.**

```{r}
month_group <- uber_2014 %>%
          group_by(month) %>%
             dplyr::summarize(Total = n()) 
datatable(month_group)
```

```{r}
ggplot(month_group , aes(month, Total, fill = month)) + 
        geom_bar( stat = "identity") +
           ggtitle("Trips by Month") +
            theme(legend.position = "none") +
            scale_y_continuous(labels = comma) +
            scale_fill_manual(values = colors)
```

```{r}
month_weekday <- uber_2014 %>%
         group_by(month, dayofweek) %>%
             dplyr::summarize(Total = n())

ggplot(month_weekday, aes(month, Total, fill = dayofweek)) + 
       geom_bar( stat = "identity", position = "dodge") +
          ggtitle("Trips by Day and Month") +
           scale_y_continuous(labels = comma) +
           scale_fill_manual(values = colors)
```


#### Finding out the number of Trips by bases

**In the following visualization, we plot the number of trips that have been taken by the passengers from each of the bases. There are five bases in all out of which, we observe that B02617 had the highest number of trips. Furthermore, this base had the highest number of trips in the month B02617. Thursday observed highest trips in the three bases - B02598, B02617, B02682.**

```{r}
ggplot(uber_2014, aes(Base)) + 
 geom_bar(fill = "darkred") +
 scale_y_continuous(labels = comma) +
 ggtitle("Trips by Bases")
```

```{r}
ggplot(uber_2014, aes(Base, fill = month)) + 
 geom_bar(position = "dodge") +
 scale_y_continuous(labels = comma) +
 ggtitle("Trips by Bases and Month") +
 scale_fill_manual(values = colors)
```

```{r}
ggplot(uber_2014, aes(Base, fill = dayofweek)) + 
 geom_bar(position = "dodge") +
 scale_y_continuous(labels = comma) +
 ggtitle("Trips by Bases and DayofWeek") +
 scale_fill_manual(values = colors)
```

#### Creating a Heatmap visualization of day, hour and month

**In this section, we will learn how to plot heatmaps using ggplot(). We will plot five heatmap plots -**


**First, we will plot Heatmap by Hour and Day.**

**Second, we will plot Heatmap by Month and Day.**

**Third, a Heatmap by Month and Day of the Week.**

**Fourth, a Heatmap that delineates Month and Bases.**

**Finally, we will plot the heatmap, by bases and day of the week.**

```{r}
day_a_hour <- uber_2014 %>%
         group_by(day, hour) %>%
            dplyr::summarize(Total = n())
datatable(day_a_hour)
```

```{r}
ggplot(day_a_hour, aes(day, hour, fill = Total)) +
            geom_tile(color = "white") +
              ggtitle("Heat Map by Hour and Day")
```

```{r}
ggplot(day_month_group, aes(day, month, fill = Total)) +
            geom_tile(color = "white") +
              ggtitle("Heat Map by Month and Day")
```

```{r}
ggplot(month_weekday, aes(dayofweek, month, fill = Total)) +
            geom_tile(color = "white") +
              ggtitle("Heat Map by Month and Day of Week")
```

```{r}
month_base <-  uber_2014 %>%
                    group_by(Base, month) %>%
                        dplyr::summarize(Total = n()) 

day0fweek_bases <-  uber_2014 %>%
                    group_by(Base, dayofweek) %>%
                        dplyr::summarize(Total = n()) 

ggplot(month_base, aes(Base, month, fill = Total)) +
            geom_tile(color = "white") +
              ggtitle("Heat Map by Month and Bases")
```

```{r}
ggplot(day0fweek_bases, aes(Base, dayofweek, fill = Total)) +
            geom_tile(color = "white") +
              ggtitle("Heat Map by Bases and Day of Week")
```


#### Creating a map visualization of rides in New York

**In the final section, we will visualize the rides in New York city by creating a geo-plot that will help us to visualize the rides during 2014 (Apr - Sep) and by the bases in the same period.**

```{r}
min_lat <- 40.5774
max_lat <- 40.9176
min_long <- -74.15
max_long <- -73.7004


ggplot(uber_2014, aes(x=Lon, y=Lat)) +
  geom_point(size=1, color = "orange") +
     scale_x_continuous(limits=c(min_long, max_long)) +
      scale_y_continuous(limits=c(min_lat, max_lat)) +
        theme_map() +
           ggtitle("NYC MAP BASED ON UBER RIDES DURING 2014 (APR-SEP)")

ggplot(uber_2014, aes(x=Lon, y=Lat, color = Base)) +
  geom_point(size=1) +
     scale_x_continuous(limits=c(min_long, max_long)) +
      scale_y_continuous(limits=c(min_lat, max_lat)) +
       theme_map() +
          ggtitle("NYC MAP BASED ON UBER RIDES DURING 2014 (APR-SEP) by BASE")
```

### Summary

**In the Uber data analysis project, We made use of packages like ggplot2 that allowed us to plot various types of visualizations that pertained to several time-frames of the year. With this, we could conclude how time affected customer trips. Finally, we made a geo plot of New York that provided us with the details of how various users made trips from different bases.**